---
title: "2020 Vote Counts: A Simple Analysis"
author: Bill Simons
output: html_document
date: "2023-06-17"
---
[Source repository](https://github.com/bill-simons/data-analysis/tree/main/2020Votes)


## State Vote Totals for 2020 US Presidential Election

I have heard a couple of times recently that, in the 2020 presidential election, Trump received more votes in California than in New York and Biden received more votes in Florida than in Massachusetts. This makes sense for a close election: the most populous states will have the most votes for each of the leading candidates. Here I will make a couple of plots to show this phenomenon.

```{r initialization, message=FALSE, warning=FALSE, include=TRUE}
library(tidyverse)
library(ggthemes)
library(patchwork)

my_theme <- function(...) { 
  theme_wsj(...) +
  theme(title=element_text(size=12))  
}
```
Vote totals were extracted from [https://www.fec.gov/resources/cms-content/documents/2020presgeresults.pdf](https://www.fec.gov/resources/cms-content/documents/2020presgeresults.pdf) using the [Tabula](https://tabula.technology/) utility program and consolidated into a single CSV file. Votes counts for irrelevant candidates were omitted from the csv file.

```{r load_data}
votes <- read.csv("2020PresResults.csv")
votes <- as_tibble(votes)

# Replace NA with zero
# Add a column for the Electoral vote winner for each state (candidate with most EC votes in the state)
# For Biden and Trump: re-scale vote counts to millions of votes, and add "fraction of votes" column
votes <- votes %>% 
         mutate_all(~replace(., is.na(.), 0)) %>%
         mutate(EV_WINNER=factor(ifelse(EV_BIDEN > EV_TRUMP,"BIDEN","TRUMP")),
                BIDEN_MILLIONS=BIDEN/1e6,
                BIDEN_PCT=BIDEN/TOTAL_VOTES,
                TRUMP_MILLIONS=TRUMP/1e6,
                TRUMP_PCT=TRUMP/TOTAL_VOTES)
print(votes)
```

First let's just do a data sanity check by taking a quick peek at it...

```{r number_of_states_won, fig.show="hold", out.width="50%"}
ggplot(votes) + 
  geom_bar(aes(y=EV_WINNER),width=0.5) + 
  ggtitle("Number of States Won",subtitle="Candidate with Most EC Votes in a State") +
  xlab("") + my_theme()

evtibble = tibble(EV=c(sum(votes$EV_BIDEN),sum(votes$EV_TRUMP)),
  CANDIDATE=c("BIDEN","TRUMP"))

ggplot(evtibble) +
  geom_col(aes(x=EV,y=CANDIDATE),width=0.5) + 
  ggtitle("Number of Electoral College Votes") +
  xlab("") +  my_theme()

```

And also take a look at the relative performance of the two candidates in each of the states.

```{r boxplots, fig.show="hold", out.width="50%"}
biden_votes <- votes %>% 
  select("STATE","BIDEN_MILLIONS","BIDEN_PCT") %>% 
  mutate(CANDIDATE="BIDEN") %>% 
  rename(MILLIONS=BIDEN_MILLIONS,PCT=BIDEN_PCT)

trump_votes <- votes %>% 
  select("STATE","TRUMP_MILLIONS","TRUMP_PCT") %>% 
  mutate(CANDIDATE="TRUMP") %>% 
  rename(MILLIONS=TRUMP_MILLIONS,PCT=TRUMP_PCT)

bycandidate <- bind_rows(biden_votes,trump_votes)

ggplot(bycandidate) +
  geom_boxplot(aes(MILLIONS,CANDIDATE)) + 
  ggtitle("Millions of Votes By State") + 
  my_theme()

ggplot(bycandidate) +
  geom_boxplot(aes(PCT,CANDIDATE)) + 
  ggtitle("Fraction of Votes Cast By State") + 
  my_theme()
```



And now let's get to the original question: in which states did each candidate get the most votes, regardless of whether they won or lost the state?

```{r vote_counts_by_state, fig.height=10, fig.width=10}

trump <- votes %>% 
  arrange(TRUMP) %>% 
  mutate(MILLIONS=TRUMP_MILLIONS,
    STATE=factor(STATE,levels=unique(STATE)))  # lock-in the re-arranged order of states 

biden <- votes %>% 
  arrange(BIDEN) %>% 
  mutate(MILLIONS=BIDEN_MILLIONS,
    STATE=factor(STATE,levels=unique(STATE))) # lock-in the re-arranged order of states

# make sure the vote totals (millions) in each data frame are scaled to the same color value
# Biden per-state vote totals go up to 11.11 million, Trump max is 6.0 million
mycolorscaler <- function(x,to,from,...) {
  scales::rescale(x,to=c(0,12),from=c(0,11.11),...)
}

# Palettes that I like with the vote share feature (not sure if it's a sequential or 
# diverging scale since it looks best when the mid-scale--0.5--is a separate color?)
# 
# Purple-Green, Red-Green
# La Jolla, RdBu
# Oslo, BurgYl, RdGy, YlGn, OrRd, YlOrRd
# Reds, Oranges, Purples, Greens, Purples 2, Greens 2, Blues 3, Reds 2
# Grays, Light Grays, Teal, Burg, Red-Yellow, PuBuGn, YlOrBr
# Cork, Vik, Mako, Blue-Red 3, Purple-Blue, PRGn, BluGrn

myplot <- function(data,title) {
  mypalette <- "Purple-Green"
  myrev=F
  bigstates = data %>% filter(MILLIONS >= 2.2)
  labels = formatC(bigstates$MILLIONS, digits = 1, format = "f")
  
  myplot <- ggplot(data,height=6) + 
      geom_col(aes(MILLIONS, STATE, fill=BIDEN_PCT),linewidth=5) + 
      # geom_text(aes(x=MILLIONS-0.5,y=STATE,label=labels),data=bigstates,size=4) + 
      scale_fill_gradientn(name="Biden vote share",
                       colors=hcl.colors(n=20,palette=mypalette,rev=myrev),
                       rescaler=mycolorscaler) +
      scale_x_continuous(breaks=seq(0,10,2), 
                         minor_breaks=seq(1,11,2)) +
      coord_cartesian(xlim=c(0,11)) +
      ggtitle(title) +  xlab("Number of Votes (millions)") +  ylab("") + 
      theme_minimal() + 
      theme(legend.position="none",
            axis.text=element_text(size=11.5),
            panel.background = element_rect(fill = "#fffff8",color="#ffffff"))
  myplot
}

tplot <- myplot(trump,"Trump") + theme(plot.margin = margin(t=3,1,1,1, "lines"),
                                       legend.title = element_text(size=12),
                                       legend.text = element_text(size=10),
                                       legend.direction = "horizontal", 
                                       legend.position = c(0.85,1.07))
bplot <- myplot(biden,"Biden") 

# using `patchwork` package to perform the plot layout here
tplot + bplot + 
  plot_annotation(
    title="2020 US Presidential Election, State-Level Votes, Colored by Vote Share",
    subtitle="California provided Trump with his most number of votes. Biden received his second-most in Florida")


```